---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(learnr)
library(ggplot2)
library(palmerpenguins)

```


## Introduction

This module explores how R data object types impact our ability to use functions and accomplish tasks like creating data visualizations. The `penguins` dataset from the `{palmerpenguins}` package will be used to illustrate our examples. 


## Viewing data

When we create plots using `ggplot2`, we have to specify the name of data object and the variables. Object names in R are case sensitive, which means that the pattern of lower and uppercase letters must match exactly.  

```{r, echo=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm,
                 shape = species))
```




```{r quiz-rhetorical-q}
quiz(
  caption = "Consider",
  
  question(
    "What if we do not remember the names of all of the variables in `penguins`? Select all options we could use to learn more about the variables in `penguins`",
    answer("All hope is lost."),
    answer("We can run the name of a data object to see a printout.", correct = TRUE),
    answer("We can use functions to print variable names.", correct = TRUE),
    answer(
      "We can type `the data object's name`penguins$` in the console and hit tab to show a list of variable names.",
      correct = TRUE
    ),
           incorrect = "It is okay if you missed some of the methods we can use to view variables. We will review them in this module."
  )
)
```



## Summary view

Simply by running the data object name as code, we can print out a summary of the data.

```{r basic-summary, exercise = TRUE}
penguins
```

This view is helpful, but it may be insufficient for data sets with large numbers of columns. 



```{r quiz-summary-view}
quiz(
  caption = "Take a Closer Look",
  
  question("Describe the kind of information we can see by simply running a data object's name. (NB: The tutorial will show you the correct answer, but you have to compare your answer to the key.)",
           type = "learnr_text", 
           answer("You can see the dataset dimensions (i.e., number of rows and columns), data values, variable names, variable types. Because you see data values, you might see potential data issues like missing values (denoted as NA).", correct = TRUE),
           incorrect = "Review your answer. Did you mention everything important? Did you notice a detail not included in the answer key?", 
           message = "You can see 1) the dataset dimensions (i.e., number of rows and columns), 2) data values, 3) variable names, 4) variable types. Because you see data values, you might see 5) data issues like missing values (denoted as NA).")
)
```




### List of variable names

If we only need a list of variable names, then we can use `names()` or `colnames()`.

```{r view-names, exercise=TRUE}
names(penguins)

```



## Detailed summary

Perhaps we want a more detailed summary of our data. We can use `str()` (short for _structure_) to see the name and type of all variables in our data set. Being able to view the **variable name and type** is especially useful for data visualization because our plot choice will depend on the data type(s) we want to see. 

### Check your understanding

```{r quiz}
quiz(
  question("Select all of the variable names below that represent `factor` type variables in the penguins dataset. You can use the code area below to apply the functions described in this module.",
    answer("species", correct = TRUE),
    answer("island" , correct = TRUE),
    answer("body_mass_g"),
    answer("year")
  )
)
```

```{r blank-exercise, exercise = TRUE}

```



## Determining data object type

R contains many data object types, including lists, arrays, matrices, vectors, and **data.frames**. Data.frames are the most commonly used type of data object in R. They are two-dimensional objects, meaning that to identify a specific value in the data set, you need to know two pieces of information: its row and its column position. Data.frames are popular because they provide a logical way to group related variables together.

Let's look at an example other than `penguins`. 

Run the following code chunk.
```{r examine-mtcars, exercise=TRUE}
mtcars
```

When we run the code `mtcars`, R outputs a rectangular grid of values. In this case, each row represents one type of car, and each column contains data about different aspects of the cars.  


We can confirm that `mtcars` is a data.frame using the `class` function.

```{r class-mtcars, exercise=TRUE}
class(mtcars)
```

It is important to verify data object type because data types impact the ways we can access information stored in the data object. We will spend lots of time learning about how to access and extract information from data objects. For now, just know that the object type will determine the kinds of functions and extract methods we can use.


## Data column types

Each of the columns in a data.frame represents a single type of data. The simplest data types are numeric/integer, character, and logical. You can determine a data column's type using the same `class()` function we used to verify object type for a data.frame. 

To let R know that you want the class of a column and not the whole data object, we use a dollar sign (`$`) to connect the data object's name with the column's name.

Check it out:
```{r column-class, exercise=TRUE}
class(penguins$species)

class(penguins$sex)

class(penguins$body_mass_g)
```


## An important data.frame feature

Print out the `mtcars` object. Scroll through the pages and look at all of the values. 

```{r examine-mtcars2, exercise=TRUE}
mtcars
```


```{r quiz-rect-data}
quiz(question(
  "What do you notice about the number of values in each column?",
  answer("All columns have the same number of values (i.e., rows)", correct = TRUE),
  answer("All columns have a different number of values (i.e., rows)"), 
  incorrect = "Are you sure? Look more closely.",
  message = "Exactly! All columns have the same number of values."
))
```

### Verifying column size

Every column within a data.frame has values in the same number of rows as every other column. You can prove that all of the variables (i.e., columns) in `mtcars` have the same number of observations (i.e., rows) using the `length()` function.

```{r examine-mtcars-col-length, exercise=TRUE}

# For example:
length(mtcars$mpg)
length(mtcars$cyl)
length(mtcars$disp) 
```

We can verify that `mpg`, `cyl`, and `disp` all have 32 values. Even if we had missing data, we would still include a symbol (NA) in the dataset to indicate that the measurement of that variable for that observation is not available.


## Check your understanding

Use what you have learned to view the `mtcars` and `penguins` datasets and answer the following questions.

```{r blank-exercise-mtcars, exercise = TRUE, exercise.lines = 5}


```

```{r blank-exercise-mtcars-hint}
# The most helpful functions are the ones that show
# both variable name and type.

```



```{r quiz2}
quiz(
  question("How many variables are in `mtcars`?",
    answer("5"),
    answer("8"),
    answer("11", correct = TRUE),
    answer("32")
  ),
  question("Which of the following are variable names in `mtcars`?",
    answer("engine"),
    answer("mpg", correct = TRUE),
    answer("axle"),
    answer("gear", correct = TRUE)
  ), 
  
  question("How many variables are in `penguins`?",
    answer("8", correct = TRUE),
    answer("10"),
    answer("35"),
    answer("344"),
    incorrect = "Variables are typically stored as columns in data.frames. Verify that you are counting columns (not rows) and try again.",
    allow_retry = TRUE
  ),
  question("Which function(s) will show variable names, variable types, data object dimensions, and actual data values for the first few observations? Select all of the functions that fit these criteria. (You can use the ? operator to pull up help pages.)",
    answer("str()", correct = TRUE),
    answer("dim()"),
    answer("head()"),
    answer("tail()")
  )
)
```



