---
title: "Tutorial"
output: 
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)

library(ggplot2)
library(palmerpenguins)
```


## Hello, World!

Welcome to R programming. This document is a `learnr` module designed to introduce you to basic concepts and constructs that are part of the R language. During this module, you will learn how to create data objects, determine a data object's type, and use functions to manipulate data objects. We also will explore examples illustrating the utility of R for research management and data handling.

## Code Chunks 

Throughout this module, you will see special areas where you can execute R code. You should use these areas to experiment and to complete tasks that check your understanding. **Important** These coding areas are not good for note-taking. Whatever you put in there will get erased eventually, so be sure to take notes somewhere else.

### Running Code

Click `Run Code` to run the entire code chunk
```{r two-plus-two, exercise=TRUE}
2 + 2

40 / 8

7 - 15
```

You should see three outputs: 4, 5, and -8. These outputs match the three operations in the above code chunk. You can clear the outputs by clicking `Start Over`. Try writing your own mathematical operations in the code area. Click `Run Code` when you are ready to see if your code can execute


### Hints and Solutions

Some of the exercises will have hints and/or solutions to guide you. You can see the information by clicking on the corresponding buttons. If the problem contains multiple hints, you can scroll through them one at a time. You will know multiple hints are available if you see **Hints** rather than **Hint**. If a problem has both a hint (or hints) and a solution, the last hint will actually be the solution

Task: Modify the code below so that only the first 3 rows of the `mtcars` dataset are shown.

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

```{r print-limit-solution}
head(mtcars, n = 3) # correct answer
```


### Quizzes

You also will encounter short quizzes. Like the code exercises, the quizzes are for your benefit. They are designed to highlight critical concepts and provide feedback to you in real-time as you work through the material.

```{r quiz}
quiz(
  question("True or False: I should look at the hints and solutions for code exercises rather than trying to work them myself first.",
    answer("True"),
    answer("False" , correct = TRUE)
  ),
  question("If you have questions about this learnr module or any others, what can you do to get help, feedback, or clarity? Select all that apply.",
    answer("Work the problems with classmates", correct = TRUE),
    answer("Search online", correct = TRUE),
    answer("Email the TA", correct = TRUE),
    answer("Email the instructor", correct = TRUE)
  )
)
```


## Learning Through Visualization

The ability to produce high quality data visualizations is one of the biggest advantages of R. Conveniently, data visualization is an excellent mechanism for learning R. 

For the rest of the module, we will cover R language basics in the context of data visualization. 


## Penguin Time

Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER) Program traveled to Antarctica to collect data about the penguin populations living in the Palmer Archipelago. The data is available to us thanks to that team and Allison Horst through the `palmerpenguin` package. Thousands of R packages (also called libraries) are available, and they vary widely in their purpose and usefulness. 

The `palmerpenguin` package is a data package. You can learn more about the `palmerpenguin` package by running the code `?palmerpenguins`.

```{r package-question, exercise = TRUE}
?palmerpenguins
```


### Viewing Data
There are two datasets in the `palmerpenguin` package: `penguins` and `penguins_raw`. We can view each data set by running the data set's name. 

Click *Run Code* to see the data set summary output.
```{r view-data, exercise = TRUE}
penguins
```

The data set summary tells us the number of rows and columns in the data set, and it shows us first ten values of a subset of columns. At the top of each column is a name (e.g., **species**) and the data type contained in each column. The exact number of columns visible will depend on your screen and text size. 

```{r summary-quiz}
quiz(
  question("How many columns are in the penguins dataset?",
    answer("5"),
    answer("8" , correct = TRUE),
    answer("10"),
    answer("344")
  )
)
```


## Exploring Data
Let's say we wanted to know more about the data in `penguins`. Specifically, let's say that we want to get an idea of the relationship between bill depth and bill depth. We could scroll through the data set printout and look at the values, but that strategy is a bad idea for many reasons. Not only would it be extremely time consuming to scroll through each row of the data set, but also we could not easily build on what we had seen. What if we next want to know if relationship between bill depth and bill depth is the same across penguin species. Are we really going to scroll through the data set all over again?

Of course not. We are going to use R to look at the data. 
```{r penguin-plot, echo=TRUE, message=FALSE}
# This code creates a scatter plot
 
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm))
```


The plot above is a scatter plot that was created using the `ggplot2` package. The `ggplot2` package is a fantastic resource for data visualization, and it builds plots using information that we supply.

Let's look at the first line:

```
ggplot(data = penguins) +
```

We know `penguins` is our data object. We looked at it earlier. `ggplot()` is a function. R functions allow you to interact with data. In our case, we want to see our data graphically. You can identify functions easily in R because they almost always are followed by parentheses (e.g., `ggplot(`...`)`) 

The `+` sign tells R that we have more information to give it. 


```{r ggplot-quiz}
quiz(
  question("The next line of code starts with `geom_point(...`. What do you think `geom_point()` is?",
    answer("A data object like `penguins`"),
    answer("A function" , correct = TRUE)
  )
)
```


```
geom_point(aes(x = bill_length_mm, y = bill_depth_mm))
```

### Aesthetics

The next line of code starts with two functions. `geom_point()`tells R that we want a scatter plot and `aes()` tells R that we want to map variables onto the scatter plot. You can remember that `aes()` impacts the way your figure looks by remembering that "aes" is short for _aesthetics_. 

Next, we tell R that the variables we want to visualize are `bill_length_mm` and `bill_depth_mm`. These variable names come directly from the column names in `penguins`. R knows to look for these variables in the `penguins` data set because we told R in first line that `data = penguins`. 

If you try to visualize a variable that is not in `penguins` or you misspell a variable's name, the plot will not appear. 

Below, `bill_length_mm` has been written incorrectly. Variable names are case sensitive, which means that the pattern of lower and uppercase letters must match exactly to be considered spelled correctly.

What happens if a misspelled variable is mapped onto the x axis?
```{r mistake-plot, exercise=TRUE}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm))

```

```{r ggplot-error-quiz}
quiz(
  question("The next line of code starts with `geom_point(...`. What do you think `geom_point` is?",
    answer("object 'Bill_Length_mm' not found", correct = TRUE),
    answer("Warning: R could not plot."),
    answer("No plot or message appears.")
  )
)
```


## Exploration Continues

We have seen how to create one specific figure, but what if we want to change the figure? What if we want to see how penguin species relates to bill length and depth?

```{r penguin-plot2, echo=TRUE, message=FALSE}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species))
```

We can see the same plot as before except now each point has been colored to indicate the species of penguin. A legend has appeared on the right side of the plot to show the mapping between each color and each species name in the `species` variable.

### Designing for Accessibility

We must be diligent during our figure design to select features that maximize the accessibility, interpretability, and accuracy our work. One of the most important considerations is the use of color. Approximately 8% of the world's population has difficulty distinguishing colors, and it is critical to realize that color vision deficiency is not limited to red-green color blindness. 

The default color options in `ggplot2` are not universally accessible. We could download another R package like `colorbrewer` that contains color palettes explicitly designed to maximize accessibility. For now, let's keep our figure monochromatic (i.e., black, greys, and white) and find a way still to distinguish the points by species using another aesthetic.   


```{r point-aesthetics-quiz}
quiz(
  question("What figure aesthetics could we use to distinguish species? Select all that apply.",
    answer("Size", correct = TRUE),
    answer("Transparency", correct = TRUE),
    answer("Shape", correct = TRUE)
  ),
  question("What would be the best aesthetic option for distinguishing the points by species?",
    answer("Size"),
    answer("Transparency"),
    answer("Shape", correct = TRUE),
    answer("All options are equally good.")
  )
)
```


Species is a categorical variable. The best option for representing different categorical variables is to use distinct colors or shape. Size is problematic to use because that aesthetic choice would suggest to the readers that the point size differences were relevant to species differences, which is not the case. Transparency is not a good option because that aesthetic exists on a continuous scale (0-100% transparent). For categorical variables, we want to use aesthetics with distinct differences that do not imply an order to the variable levels.

### Exercise 1

Modify the code below so that species is mapped on to `shape` rather than `color`. 

```{r penguin-plot3, exercise = TRUE}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species))
```


```{r penguin-plot3-solution}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 shape = species))
```


### Exercise 2 

Modify this plot so that both point size and point color are mapped to `species`.
```{r penguin-plot4, exercise = TRUE}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species))
```

```{r penguin-plot4-solution}

ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species, shape = species))
```


## Help!

If you are nervous because you are worried that you will have to memorize thousands of keywords and code lines to make simple plots, never fear: programming requires almost no memorization.


### R Help Pages

Programmers use a variety of tools to remind themselves which functions to use and what arguments each function takes. We can see the specific arguments list for any function by running the code `?<functionName>` where `<functionName>` is the name of a function (do not include parentheses). 

Run the following code to see the R help page for `geom_point()`.
```{r geom_point-help, exercise = TRUE}
?geom_point
```

Scroll all the way down to "Aesthetics". You should see "`geom_point()` understands the following aesthetics (required aesthetics are in bold):" followed by a list of letters and words. Each word or letter (e.g., x, y, alpha) is an argument that `geom_point()` understands and can use to change how the plot looks. 


### Cheatsheets   
Some packages have cheatsheets to guide users. The `tidyverse` is a collection of packages that work well together. Many of these packages, including `ggplot2`, have fantastic cheatsheets. You can find the entire collection of tidyverse cheatsheets online (https://www.rstudio.com/resources/cheatsheets). The sheets can be intimidating at first, but learning how to use them is critical for your development as an R user and researcher.

We will spend a lot of time practicing how to use cheatsheets and R documentation to find information and to troubleshoot coding problems.

### StackOverflow and Online Forums
StackOverflow.com is a great resource for programmers.

## Next Steps

Congratulations, team! You have embarked on your first R adventure and lived to tell the tale of penguins, plots, and problem solving strategies. Next, we will take a deeper look at R data types.

## Additional Practice

### Exercise 1

Create a scatter plot comparing bill length (`bill_length_mm`) and flipper length (`flipper_length`). 

```{r add-exercise-1, exercise=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = ... , y = ...))
```

```{r add-exercise-1-solution}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = flipper_length))
```


### Exercise 2

Modify this scatter plot so that the point shape corresponds to penguin sex (`sex`).


```{r add-exercise-2, exercise=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g))
```

```{r add-exercise-2-solution}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g,
                 shape = sex))
```


### Exercise 3

By default, all points in a scatter plot have the size of 1 unit. Modify the following plot so that all points are `size = 2`. This kind of aesthetic is called a _set_ aesthetic because it is not tied to a variable--it is constant for all values. Notice that set aesthetics do not go inside the `aes()` function.

```{r add-exercise-3, exercise=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g), 
             size = 1)
```

```{r add-exercise-3-solution}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g),
             size = 2)
```



### Exercise 4

In this figure, someone has attempted to set all of the points to be the color **blue**, but something has gone very wrong. All of the points are red and there is now a legend that says red points are blue. How can we fix this code?

```{r add-exercise-4, exercise=TRUE}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g))
```

```{r add-exercise-4-hint}
# The code color = "blue" is written correctly, but it is in the wrong place. Where should it be moved? 
```

```{r add-exercise-4-solution}
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm , y = body_mass_g, 
                 color = "blue"))
```


### Exercise 5

Modify this density plot by lowering the `alpha`. Alpha is a measure of transparency where 0 = fully transparent and 1 = fully opaque. Place the alpha argument within the `geom_density()` function but outside of the `aes()` function.

```{r add-exercise-5, exercise=TRUE}
ggplot(data = penguins) +
  geom_density(aes(bill_length_mm, 
                   fill = species))
```


```{r add-exercise-5-hint}
# Try selecting an alpha value between 0.30 and .70
```

```{r add-exercise-5-solution}
ggplot(data = penguins) +
  geom_density(aes(bill_length_mm, 
                   fill = species),
                   alpha = 0.2) # any value between ~0.30 and ~.70 is interpretable
```



```{r}

# Pre-class assessment questions
# 
# What package was used for plotting?
# A) ggplot2
# B) geom_point()
# C) ggplot()
# D) penguins
# 
# 
# What function creates a scatter plot?
# A) ggplot()
# B) geom_point()
# C) penguins()
# D) scatter()
# 
# 
# How can you identify a function?
# A) Functions are followed by curly brackets.
# B) Functions are followed by square brackets.
# C) Functions are followed by an equals sign.
# D) Functions are followed by parentheses.
# 
# 
# Which exercise in the Additional Exercises section was the hardest for you?
# A) 1
# B) 2
# C) 3
# D) 4
# E) 5
# 
#
# What questions do you have from this tutorial?
# 
# 
# Is there anything specific you would like to review or practice in class? Note: "All of it." is not an appropriate answer because it is not specific.


```
